Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-11-07 rnni_convexity.tex
pass to first mention of Bubble Sort

We are now ready to introduce $\findpath$ (Algorithm~\ref{alg:find_path}), an efficient algorithm for computing paths between any two given trees $T$ and $R$ in $\rnni$.
Let $[C_1, \ldots, C_{n-2}]$ be the cluster representation of $R$ (the ``destination'' tree).
Then $C_i$ is induced by the node of rank $i$ for $i = 1, \ldots, n-2$.
$\findpath$ computes a path $p$ which initially consists of $T$ only and is extended stepwise.
In order to do this, the clusters $C_1, \ldots, C_{n-2}$ of $R$ are considered in their order.
Let $T'$ be the last tree of $p$ when arriving at cluster $C_k$.
We assume that the clusters $C_1, \ldots, C_{k-1}$ are induced by nodes of rank $1, \ldots, k-1$ in $T'$, respectively.
In the first step it is $T' = T$ and $C_k = C_1$.
The aim is now to find a sequence of $\rnni$ moves, starting at $T'$, that do not change the clusters $C_1, \ldots, C_{k-1}$, but result in a tree containing clusters $C_1, \ldots, C_{k}$.
Therefore, we need to find the smallest cluster containing all elements of $C_k$ in $T'$.
This node is denoted by $\mrca_{T'}(C_k)$ (short for \emph{most recent common ancestor}).
The rank of $\mrca_{T'}(C_k)$ in $T'$ is higher than the rank of the node inducing $C_k$ in $R$, because $C_1, \ldots, C_{k-1}$ are already present in $T'$ and we assume that $C_k$ is not.
Now perform an $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$.
The existence of such a move is not obvious but we will see later (Lemma~\ref{lemma:mrca_move}) that this kind of move always exists as long as $T'$ does not induce the cluster $C_k$.
So we can perform such moves until a tree $T''$ is reached with $\mrca_{T''}(C_k)$ inducing $C_k$ and having has the same rank $k$ in $T''$ as in $R$.
Appending all trees on this path from $T'$ to $T''$ and repeating this procedure for all clusters of $R$ results in a path $p$ from $T$ to $R$.

\begin{algorithm}[H]
\caption{$\findpath$($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
\STATE $T' := T$
\STATE $p := [T']$
\FOR {$i = 1, \dots, n-2$}
\STATE Let $C_i$ be the cluster induced by the node with rank $i$ in $R$ \label{alg:find_path:line:cluster}
\WHILE {$\rank_{T'}(\mrca_{T'}(C_i))>i$}
\STATE Update $T'$: Decrease $\rank_{T'}(\mrca_{T'}(C_i))$ by an $\rnni$ move \label{alg:findpath:line:move_set_down}
\STATE $p = p+T'$
\ENDWHILE
\ENDFOR
\RETURN $p$
\end{algorithmic}
\end{algorithm}

@@

\begin{lemma}
Let $T$ and $R = [C_1, \ldots, C_{n-2}]$ be two trees and $p = \findpath (T,R)$.
Let $T'$ be a tree on $p$ different from both $T$ and $R$ such that $C_1, \ldots, C_{k-1}$ are induced by internal nodes of $T'$, but $C_k$ is not.
Then there exists exactly one $\rnni$ move on $T'$ that results in a tree $T''$ such that
\todo{What's $i$?}
$\mrca_{T''}(C_i) < \mrca_{T'}(C_i)$.
\label{lemma:mrca_move}
\end{lemma}

@@

With Lemma~\ref{lemma:mrca_move} we can conclude that $\findpath$ is a deterministic algorithm.
Its running time is quadratic in the number of taxa.
Since it computes a path between two arbitrary trees in $\rnni$, we can use the length of such a path as approximation for the distance between the trees.
A natural question is how well this estimates the actual distance in $\rnni$.
To answer this question for small trees, we implemented $\findpath$.
\todo{LC: github - link}
For computing the actual $\rnni$ distances we used the algorithm for computing the $\rnni$ graph as described in Section~\ref{section:alg_RNNI_graph}.

@@

For introducing an algorithm to compute paths between caterpillar trees, we need further notations.
In this section we aim to investigate properties of paths between \emph{caterpillar trees} where each internal node has at least one child that is a leaf.
A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called \emph{caterpillar path}.
If $p$ is shortest among all caterpillar paths between these trees we call it \emph{shortest caterpillar path}, and the length of $p$ will be denoted by $d_c(T,R)$ and called \emph{caterpillar distance}.

Let us now introduce a representation of caterpillar trees as lists of taxa.
This representation is basically a simplification of the cluster representation for caterpillar trees and works as follows.
For a caterpillar tree with cluster representation $T = [C_1, \ldots, C_{n-2}]$ it is $C_i \subset C_{i+1}$ with $|C_i| = |C_{i+1}|$ for all $i = 1, \ldots, n-3$.
Therefore, the list of taxa $L_T = [x_1, x_2, \ldots, x_n]$ with $x_1<x_2$ uniquely represents the caterpillar tree $T = [C_1, \ldots, C_{n-2}]$ with $C_i = \{x_1, \ldots, x_{i+1}\}$.
In this list representing a caterpillar tree the taxa are ordered non-decreasing, according to the ranks of their parents.
Note that the only two elements in such a list that have parents of equal rank are the first two elements $x_1$ and $x_2$.
That is why we order them in the list such that $x_1 < x_2$ to make the list representation of caterpillar trees unique.
We will call a pair of two taxa that share their parent \emph{cherry}.
We will use this list representation to show how one can adapt the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.

We are now going to explain what kind of $\rnni$ moves are possible between caterpillar trees and see how such a move changes the list representation of a tree.
At first it is important to realise that the only possible $\rnni$ moves between caterpillar trees are $\nni$ moves, since there is no pair of internal nodes with rank difference one that is not connected by an edge.
$\nni$ moves between caterpillar trees correspond to swaps of two neighboured elements in the list representation.
The only exception of this are $\nni$ moves on the lowest internal edge, that is the edge connecting the internal nodes of rank one and two.
There are two $\nni$ moves possible on this edge that result in caterpillar trees, while there is for every other edge only one $\nni$ move that results in a caterpillar tree.
Corresponding to the $\nni$ moves on the lowest internal edge are exchanges of either the first or the second element with the third element in the list representation.

By using this knowledge we can compute shortest caterpillar paths by modifying Bubble Sort as described in the following.
We call this algorithm $\csort$ (Algorithm~\ref{alg:csort}).

Let $T$ and $R$ be the input trees for $\csort$.
We assume that they are given in their list representation $L_T$ and $L_R$ as described above and want to compute a path from $L_T$ to $L_R$.
This path $p$ is built stepwise by adding a new list to the end of $p$.
We call the last element of $p$ at any step $L_{T'}$, representing a tree $T'$ and when we say we update $L_{T'}$, we modify this list and append the new version of $L_{T'}$ to $p$.

>>>>

\todo{I suggest to rewrite this paragraph following the standard structure:\\
\emph{Step $0$. $T_0 = T$, ...\\
Step $k+1$. $T_{k+1}$ is defined...}}
We are now ready to introduce $\findpath$ (Algorithm~\ref{alg:find_path}), an efficient algorithm for computing paths between any two given trees $T$ and $R$ in $\rnni$.
Let $[C_1, \ldots, C_{n-2}]$ be the cluster representation of $R$ (the ``destination'' tree).
Then $C_i$ is induced by the node of rank $i$ for $i = 1, \ldots, n-2$.
$\findpath$ computes a path $p$ which initially consists of $T$ only and is extended stepwise.
In order to do this, the clusters $C_1, \ldots, C_{n-2}$ of $R$ are considered in their order.
Let $T'$ be the last tree of $p$ when
\todo{This term is not clear or defined.}
arriving at cluster $C_k$.
We assume that the clusters $C_1, \ldots, C_{k-1}$ are induced by nodes of rank $1, \ldots, k-1$ in $T'$, respectively.
In the first step it is $T' = T$ and $C_k = C_1$.
The aim is now to find a sequence of $\rnni$ moves, starting at $T'$, that do not change the clusters $C_1, \ldots, C_{k-1}$, but result in a tree containing clusters $C_1, \ldots, C_{k}$.
Therefore, we need to find the smallest cluster containing all elements of $C_k$ in $T'$.
This node is denoted by $\mrca_{T'}(C_k)$ (short for \emph{most recent common ancestor}).
The rank of $\mrca_{T'}(C_k)$ in $T'$ is higher than the rank of the node inducing $C_k$ in $R$, because $C_1, \ldots, C_{k-1}$ are already present in $T'$ and we assume that $C_k$ is not.
Now perform an $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$.
\todo{I think it is obvious.
I suggest to remove this Lemma and save it somewhere in case we get an too sensitive referee.
Tag a commit?
Or branch? -- whatever works.}
The existence of such a move is not obvious but we will see later (Lemma~\ref{lemma:mrca_move}) that this kind of move always exists as long as $T'$ does not induce the cluster $C_k$.
So we can perform such moves until a tree $T''$ is reached with $\mrca_{T''}(C_k)$ inducing $C_k$ and having has the same rank $k$ in $T''$ as in $R$.
Appending all trees on this path from $T'$ to $T''$ and repeating this procedure for all clusters of $R$ results in a path $p$ from $T$ to $R$.

\todo{$p = p+T'$: Do we really want to be updating this in every while loop?}
\begin{algorithm}[H]
\caption{$\findpath$($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
\STATE $T' := T$
\STATE $p := [T']$
\FOR {$i = 1, \dots, n-2$}
\STATE Let $C_i$ be the cluster induced by the node with rank $i$ in $R$ \label{alg:find_path:line:cluster}
\WHILE {$\rank_{T'}(\mrca_{T'}(C_i))>i$}
\STATE Update $T'$: Decrease $\rank_{T'}(\mrca_{T'}(C_i))$ by an $\rnni$ move \label{alg:findpath:line:move_set_down}
\STATE $p = p+T'$
\ENDWHILE
\ENDFOR
\RETURN $p$
\end{algorithmic}
\end{algorithm}

@@

\begin{lemma}
Let $T$ and $R = [C_1, \ldots, C_{n-2}]$ be two trees and
\todo{This lemma is to prove that FindPath is an (deterministic) algorithm, so we cannot relay on this.}
$p = \findpath (T,R)$.
Let $T'$ be a tree on $p$ different from both $T$ and $R$ such that $C_1, \ldots, C_{k-1}$ are induced by internal nodes of $T'$, but $C_k$ is not.
Then there exists exactly one $\rnni$ move on $T'$ that results in a tree $T''$ such that
\todo{What's $i$?}
$\mrca_{T''}(C_i) < \mrca_{T'}(C_i)$.
\label{lemma:mrca_move}
\end{lemma}

@@

Note that the uniqueness of the $\rnni$ move in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path} implies that the algorithm is deterministic.
The running time of the algorithm is quadratic in the number of taxa.

Because the algorithm returns a path between pairs of trees in $\rnni$, the length of the path approximates the $\rnni$ distance from above.
A natural question then is how accurate this approximation is.
To answer this question for small trees, we have implemented the algorithm $\findpath$
\todo{LC: github - link}
\autocite{}
and compared the results with the true distance found using the algorithm from Section~\ref{section:alg_RNNI_graph}.

@@

In this section we aim to investigate properties of paths between \emph{caterpillar trees}, each internal node of which is adjacent to at least one leaf.
A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called a \emph{caterpillar path}.
If $p$ is the shortest pass among all caterpillar paths between $T$ and $R$ we call it a \emph{shortest caterpillar path}, and denote the length of $p$ by $d_c(T,R)$, which we call the \emph{caterpillar distance}.

Let us now introduce a representation of caterpillar trees as lists of taxa.
This representation is basically a simplification of the cluster representation for caterpillar trees and works as follows.
\todo{I don't follow these two sentences.}
For a caterpillar tree with a cluster representation $T = [C_1, \ldots, C_{n-2}]$ it is $C_i \subset C_{i+1}$ with $|C_i| = |C_{i+1}|$ for all $i = 1, \ldots, n-3$.
Therefore, the list of taxa $L_T = [x_1, x_2, \ldots, x_n]$ with $x_1<x_2$ uniquely represents the caterpillar tree $T = [C_1, \ldots, C_{n-2}]$ with $C_i = \{x_1, \ldots, x_{i+1}\}$.
In this list representing a caterpillar tree the taxa are ordered non-decreasing, according to the ranks of their parents.
Note that the only two elements in such a list that have parents of equal rank are the first two elements $x_1$ and $x_2$.
That is why we order them in the list such that
\todo{Are we assuming taxa is ordered?}
$x_1 < x_2$
to make the list representation of caterpillar trees unique.
We will call a pair of two taxa that share their parent a \emph{cherry}.

%We are now going to explain what kind of $\rnni$ moves are possible between caterpillar trees and see how such a move changes the list representation of a tree.
%At first it is important to realise that the only possible $\rnni$ moves between caterpillar trees are $\nni$ moves, since there is no pair of internal nodes with rank difference one that is not connected by an edge.
%$\nni$ moves between caterpillar trees correspond to swaps of two neighboured elements in the list representation.
%The only exception of this are $\nni$ moves on the lowest internal edge, that is the edge connecting the internal nodes of rank one and two.
%There are two $\nni$ moves possible on this edge that result in caterpillar trees, while there is for every other edge only one $\nni$ move that results in a caterpillar tree.
%Corresponding to the $\nni$ moves on the lowest internal edge are exchanges of either the first or the second element with the third element in the list representation.

The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.
Let $T$ and $R$ be the input trees given by their list representations $L_T$ and $L_R$.
This path $p$ is built stepwise by adding a new list to the end of $p$.
We call the last element of $p$ at any step $L_{T'}$, representing a tree $T'$ and when we say we update $L_{T'}$, we modify this list and append the new version of $L_{T'}$ to $p$.
